---
title: "Comparison with Alternatives"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparison with Alternatives}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5
)
library(corrselect)
```

## Overview

This vignette compares corrselect with common alternatives for handling multicollinearity and variable selection:

- **caret::findCorrelation()**: Greedy correlation-based removal
- **Boruta**: Random forest feature importance
- **glmnet**: LASSO/Ridge regularization
- **Manual VIF removal**: Iterative variance inflation factor pruning


---

## Test Dataset

We'll use the `bioclim_example` dataset with 19 WorldClim bioclimatic variables:

```{r}
data(bioclim_example)
predictors <- bioclim_example[, -1]  # Exclude response
response <- bioclim_example[, 1]

cat("Number of variables:", ncol(predictors), "\n")
cat("Sample size:", nrow(predictors), "\n")
cat("Response variable: species_richness\n")
```

Let's visualize the correlation structure:

```{r, fig.width=8, fig.height=6}
cor_matrix <- cor(predictors)

# Custom color palette
col_pal <- colorRampPalette(c("#3B4992", "white", "#EE0000"))(100)

# Correlation heatmap
par(mar = c(1, 1, 2, 1))
image(1:ncol(cor_matrix), 1:nrow(cor_matrix), t(cor_matrix[nrow(cor_matrix):1, ]),
      col = col_pal,
      xlab = "", ylab = "", axes = FALSE,
      main = "Correlation Matrix (19 Bioclimatic Variables)",
      zlim = c(-1, 1))
axis(1, at = 1:ncol(cor_matrix), labels = colnames(cor_matrix), las = 2, cex.axis = 0.7)
axis(2, at = ncol(cor_matrix):1, labels = colnames(cor_matrix), las = 2, cex.axis = 0.7)

# Add correlation values
for (i in 1:ncol(cor_matrix)) {
  for (j in 1:nrow(cor_matrix)) {
    text(i, nrow(cor_matrix) - j + 1, sprintf("%.2f", cor_matrix[j, i]), cex = 0.5)
  }
}
```


---

# Comparison 1: caret::findCorrelation()

caret's `findCorrelation()` uses a greedy heuristic to remove correlated variables.

## How It Works

```{r}
if (requireNamespace("caret", quietly = TRUE)) {
  # Run caret
  to_remove_caret <- caret::findCorrelation(cor_matrix, cutoff = 0.7)
  result_caret <- predictors[, -to_remove_caret]

  cat("Variables removed by caret:", length(to_remove_caret), "\n")
  cat("Variables kept:", ncol(result_caret), "\n")
  cat("Removed variables:", colnames(predictors)[to_remove_caret], "\n")
}
```

## corrselect::corrPrune() Comparison

```{r}
# Run corrselect (exact mode)
result_corrselect <- corrPrune(predictors, threshold = 0.7, mode = "exact")

cat("\nVariables removed by corrselect:", length(attr(result_corrselect, "removed_vars")), "\n")
cat("Variables kept:", ncol(result_corrselect), "\n")
cat("Removed variables:", attr(result_corrselect, "removed_vars"), "\n")
```

## Visual Comparison

```{r, fig.width=7, fig.height=5}
# Extract correlations
cor_orig <- cor(predictors)
cor_corrselect <- cor(result_corrselect)

if (requireNamespace("caret", quietly = TRUE)) {
  cor_caret <- cor(result_caret)

  # Overlaid histogram comparing all three
  hist(abs(cor_orig[upper.tri(cor_orig)]),
       breaks = 30,
       main = "Distribution of Absolute Correlations",
       xlab = "Absolute Correlation",
       col = rgb(0.5, 0.5, 0.5, 0.4),
       xlim = c(0, 1))

  hist(abs(cor_caret[upper.tri(cor_caret)]),
       breaks = 30,
       col = rgb(0.8, 0.2, 0.2, 0.4),
       add = TRUE)

  hist(abs(cor_corrselect[upper.tri(cor_corrselect)]),
       breaks = 30,
       col = rgb(0.2, 0.5, 0.8, 0.4),
       add = TRUE)

  abline(v = 0.7, col = "black", lwd = 2, lty = 2)
  legend("topright",
         legend = c(paste0("Original (", ncol(predictors), " vars)"),
                    paste0("caret (", ncol(result_caret), " vars)"),
                    paste0("corrselect (", ncol(result_corrselect), " vars)"),
                    "Threshold"),
         fill = c(rgb(0.5, 0.5, 0.5, 0.4), rgb(0.8, 0.2, 0.2, 0.4),
                  rgb(0.2, 0.5, 0.8, 0.4), NA),
         border = c("black", "black", "black", NA),
         lty = c(NA, NA, NA, 2),
         lwd = c(NA, NA, NA, 2),
         bty = "n")
}
```

## Key Differences

| Feature | caret::findCorrelation() | corrselect::corrPrune() |
|---------|-------------------------|------------------------|
| **Algorithm** | Greedy heuristic | Graph-theoretic (`vignette("theory")`) |
| **Reproducibility** | Non-deterministic | Fully reproducible |
| **Variables retained** | Fewer (conservative removal) | More (maximizes retention) |
| **force_in support** | No | Yes |
| **Mixed data** | No | Yes (via assocSelect) |
| **Speed** | Fast | Fast (greedy) or moderate (exact) |

**caret**: Quick exploratory analysis where reproducibility is not critical.

**corrselect**: Reproducible results and maximal variable retention.

---

## Reproducible Benchmarks

Using the `cor_example` dataset (20 variables with known block structure):

```{r}
data(cor_example)

# Ensure matrix is exactly symmetric (fix any floating point asymmetries)
cor_example <- (cor_example + t(cor_example)) / 2
diag(cor_example) <- 1

cat("Matrix dimensions:", nrow(cor_example), "x", ncol(cor_example), "\n")
cat("Mean correlation:", mean(abs(cor_example[upper.tri(cor_example)])), "\n")
```

Timing comparison at threshold = 0.7:

```{r, eval=requireNamespace("caret", quietly=TRUE) && requireNamespace("microbenchmark", quietly=TRUE)}
if (requireNamespace("caret", quietly = TRUE) &&
    requireNamespace("microbenchmark", quietly = TRUE)) {

  # Benchmark with microbenchmark
  mb <- microbenchmark::microbenchmark(
    caret = {
      to_remove <- caret::findCorrelation(cor_example, cutoff = 0.7)
      cor_example[, -to_remove]
    },
    corrselect_greedy = {
      MatSelect(cor_example, threshold = 0.7, mode = "greedy")
    },
    corrselect_exact = {
      MatSelect(cor_example, threshold = 0.7, mode = "exact")
    },
    times = 50
  )

  print(mb)
}
```

Visualization of timing results:

```{r, eval=requireNamespace("caret", quietly=TRUE) && requireNamespace("microbenchmark", quietly=TRUE), fig.width=7, fig.height=5}
if (requireNamespace("caret", quietly = TRUE) &&
    requireNamespace("microbenchmark", quietly = TRUE)) {

  # Convert to milliseconds
  mb_df <- data.frame(
    method = mb$expr,
    time_ms = mb$time / 1e6
  )

  # Compute medians for plotting
  method_levels <- c("caret", "corrselect_greedy", "corrselect_exact")
  method_labels <- c("caret", "corrselect (greedy)", "corrselect (exact)")

  medians <- sapply(method_levels, function(m) {
    median(mb_df$time_ms[mb_df$method == m])
  })

  # Barplot
  par(mar = c(8, 4, 4, 2))
  bp <- barplot(medians,
                names.arg = method_labels,
                las = 2,
                col = c(rgb(0.8, 0.2, 0.2, 0.7),
                        rgb(0.2, 0.5, 0.8, 0.7),
                        rgb(0.4, 0.4, 0.4, 0.7)),
                ylab = "Median Time (ms)",
                main = "Timing Comparison (p=20, threshold=0.7)",
                cex.names = 0.8)

  # Add median values on bars
  text(bp, medians, labels = sprintf("%.2f ms", medians),
       pos = 3, cex = 0.8)
}
```

Scaling with problem size:

```{r, eval=requireNamespace("caret", quietly=TRUE), fig.width=7, fig.height=5}
if (requireNamespace("caret", quietly = TRUE)) {

  set.seed(20250125)
  p_values <- c(10, 20, 30, 40, 50)
  times_caret <- numeric(length(p_values))
  times_corrselect_greedy <- numeric(length(p_values))
  times_corrselect_exact <- numeric(length(p_values))

  for (i in seq_along(p_values)) {
    p <- p_values[i]

    # Generate test matrix
    mat <- matrix(rnorm(p * (p + 50)), ncol = p)
    cor_mat <- cor(mat)

    # Time caret
    times_caret[i] <- system.time({
      caret::findCorrelation(cor_mat, cutoff = 0.7)
    })[3]

    # Time corrselect greedy
    times_corrselect_greedy[i] <- system.time({
      MatSelect(cor_mat, threshold = 0.7, mode = "greedy")
    })[3]

    # Time corrselect exact (skip for p > 30 to avoid long runtime)
    if (p <= 30) {
      times_corrselect_exact[i] <- system.time({
        MatSelect(cor_mat, threshold = 0.7, mode = "exact")
      })[3]
    } else {
      times_corrselect_exact[i] <- NA
    }
  }

  # Plot scaling
  plot(p_values, times_caret * 1000,
       type = "b", col = rgb(0.8, 0.2, 0.2), pch = 19, lwd = 2,
       xlab = "Number of Variables (p)",
       ylab = "Time (ms)",
       main = "Scaling with Problem Size",
       log = "y", ylim = c(0.1, max(times_corrselect_exact * 1000, na.rm = TRUE)))

  lines(p_values, times_corrselect_greedy * 1000,
        type = "b", col = rgb(0.2, 0.5, 0.8), pch = 19, lwd = 2)

  lines(p_values[!is.na(times_corrselect_exact)],
        times_corrselect_exact[!is.na(times_corrselect_exact)] * 1000,
        type = "b", col = rgb(0.4, 0.4, 0.4), pch = 19, lwd = 2)

  legend("topleft",
         legend = c("caret", "corrselect (greedy)", "corrselect (exact)"),
         col = c(rgb(0.8, 0.2, 0.2), rgb(0.2, 0.5, 0.8), rgb(0.4, 0.4, 0.4)),
         lwd = 2, pch = 19, bty = "n")
}
```

---

# Comparison 2: Boruta

Boruta identifies **important** variables via random forest permutation tests. This is fundamentally different from removing **redundant** variables.

## Different Questions

```{r, eval=requireNamespace("Boruta", quietly=TRUE)}
if (requireNamespace("Boruta", quietly = TRUE)) {
  library(Boruta)

  # Boruta: which variables predict species richness?
  set.seed(123)
  boruta_result <- Boruta(species_richness ~ ., data = bioclim_example, maxRuns = 100)

  cat("Boruta results:\n")
  print(table(boruta_result$finalDecision))

  important_vars <- names(boruta_result$finalDecision[boruta_result$finalDecision == "Confirmed"])
  cat("\nConfirmed important variables:", important_vars, "\n")
  cat("Number of important variables:", length(important_vars), "\n")
}
```

```{r}
# corrselect: which variables are redundant?
corrselect_result <- corrPrune(predictors, threshold = 0.7)
cat("\ncorrselect results:\n")
cat("Non-redundant variables:", ncol(corrselect_result), "\n")
cat("Variables:", names(corrselect_result), "\n")
```

## Comparison Table

| Aspect | Boruta | corrselect |
|--------|--------|------------|
| **Question answered** | "Which variables predict Y?" | "Which variables are redundant?" |
| **Criterion** | Variable importance | Pairwise correlation |
| **Requires response** | Yes (supervised) | No (unsupervised) |
| **Handles multicollinearity** | Partially | Yes (explicitly) |
| **Deterministic** | No (stochastic) | Yes |
| **Computational cost** | High (many random forests) | Low |

## Combined Workflow

Boruta and corrselect can be used sequentially:

```{r, eval=FALSE}
# Remove redundant predictors (corrselect)
data_pruned <- corrPrune(raw_data, threshold = 0.7)

# Identify important predictors (Boruta)
boruta_result <- Boruta(response ~ ., data = data_pruned)
final_vars <- names(boruta_result$finalDecision[boruta_result$finalDecision == "Confirmed"])

# Final model
final_model <- lm(response ~ ., data = data_pruned[, c("response", final_vars)])
```

Workflow components:

1. corrPrune reduces the variable set based on correlation structure
2. Boruta screens the reduced set for predictive importance
3. Final model contains variables that are both non-redundant and statistically important

This addresses both redundancy and variable importance when both reproducible pruning and RF-based importance tests are required.

---

# Comparison 3: glmnet (LASSO/Ridge)

glmnet performs variable selection via L1/L2 regularization, shrinking coefficients toward zero.

## How It Works

```{r, eval=requireNamespace("glmnet", quietly=TRUE)}
if (requireNamespace("glmnet", quietly = TRUE)) {
  library(glmnet)

  # Prepare data
  X <- as.matrix(predictors)
  y <- response

  # Fit LASSO with cross-validation
  set.seed(123)
  cv_lasso <- cv.glmnet(X, y, alpha = 1)

  # Extract non-zero coefficients at lambda.1se
  coef_lasso <- coef(cv_lasso, s = "lambda.1se")
  selected_lasso <- rownames(coef_lasso)[coef_lasso[, 1] != 0][-1]  # Remove intercept

  cat("glmnet selected variables:", selected_lasso, "\n")
  cat("Number of variables:", length(selected_lasso), "\n")
}
```

## Compare Prediction Performance

```{r, eval=requireNamespace("glmnet", quietly=TRUE)}
if (requireNamespace("glmnet", quietly = TRUE)) {
  # Model 1: glmnet-selected variables (OLS)
  model_glmnet <- lm(species_richness ~ .,
                     data = bioclim_example[, c("species_richness", selected_lasso)])

  # Model 2: corrselect-selected variables (OLS)
  model_corrselect <- lm(species_richness ~ .,
                         data = cbind(species_richness = response, result_corrselect))

  cat("\nPrediction performance (R²):\n")
  cat("glmnet variables:", round(summary(model_glmnet)$r.squared, 3), "\n")
  cat("corrselect variables:", round(summary(model_corrselect)$r.squared, 3), "\n")

  cat("\nNumber of predictors:\n")
  cat("glmnet:", length(selected_lasso), "\n")
  cat("corrselect:", ncol(result_corrselect), "\n")
}
```

## Visual: Coefficient Comparison

```{r, eval=requireNamespace("glmnet", quietly=TRUE), fig.width=10, fig.height=5}
if (requireNamespace("glmnet", quietly = TRUE)) {
  par(mfrow = c(1, 2), mar = c(8, 4, 3, 2))

  # glmnet coefficients (shrinkage)
  coef_vals <- coef_lasso[coef_lasso[, 1] != 0, ][-1]
  barplot(sort(abs(coef_vals), decreasing = TRUE),
          las = 2,
          main = "glmnet: Shrunk Coefficients",
          ylab = "Absolute Coefficient Value",
          col = "salmon",
          cex.names = 0.7)

  # corrselect: unbiased OLS coefficients
  coef_corrselect <- coef(model_corrselect)[-1]  # Remove intercept
  barplot(sort(abs(coef_corrselect), decreasing = TRUE),
          las = 2,
          main = "corrselect: Unbiased OLS Coefficients",
          ylab = "Absolute Coefficient Value",
          col = "lightblue",
          cex.names = 0.7)
}
```

## Key Differences

| Feature | glmnet | corrselect |
|---------|--------|------------|
| **Primary goal** | Prediction accuracy | Interpretability + multicollinearity removal |
| **Method** | Coefficient shrinkage | Hard variable removal |
| **Coefficient bias** | Yes (shrunk toward zero) | No (standard OLS) |
| **Multicollinearity** | Handles via shrinkage | Removes via pruning |
| **Requires response** | Yes | No (corrPrune) |
| **Cross-validation** | Built-in | Not needed |

**glmnet**: Prediction-focused; coefficient interpretability secondary.

**corrselect**: Interpretability focus; unbiased coefficients; exploratory analysis without response variable.

---

# Comparison 4: modelPrune() vs Manual VIF Removal

For regression models, VIF-based multicollinearity removal is common. Let's compare corrselect's `modelPrune()` with manual VIF removal.

## Manual VIF Removal (Traditional Approach)

```{r}
# Manual iterative VIF removal
manual_vif_removal <- function(formula, data, threshold = 5) {
  require(car)

  model <- lm(formula, data = data)
  vif_vals <- car::vif(model)

  iterations <- 0
  while (max(vif_vals) > threshold && iterations < 100) {
    iterations <- iterations + 1

    # Remove variable with highest VIF
    var_to_remove <- names(which.max(vif_vals))
    cat("Iteration", iterations, ": Removing", var_to_remove, "(VIF =",
        round(max(vif_vals), 2), ")\n")

    # Update formula
    formula_str <- paste(deparse(formula), collapse = "")
    formula_str <- gsub(paste0("\\+\\s*", var_to_remove), "", formula_str)
    formula_str <- gsub(paste0(var_to_remove, "\\s*\\+"), "", formula_str)
    formula <- as.formula(formula_str)

    # Refit model
    model <- lm(formula, data = data)
    vif_vals <- car::vif(model)
  }

  list(model = model, iterations = iterations, vif = vif_vals)
}

# Run manual VIF removal
if (requireNamespace("car", quietly = TRUE)) {
  cat("Manual VIF removal (iterative):\n")
  manual_result <- manual_vif_removal(species_richness ~ ., data = bioclim_example, threshold = 5)
  cat("\nFinal VIF values:\n")
  print(round(manual_result$vif, 2))
  cat("\nTotal iterations:", manual_result$iterations, "\n")
}
```

## modelPrune() Comparison

```{r}
# Run modelPrune
modelprune_result <- modelPrune(species_richness ~ ., data = bioclim_example, limit = 5)

cat("\nmodelPrune results:\n")
cat("Variables removed:", attr(modelprune_result, "removed_vars"), "\n")
cat("Variables kept:", length(attr(modelprune_result, "selected_vars")), "\n")

# Extract final model
final_model <- attr(modelprune_result, "final_model")
if (requireNamespace("car", quietly = TRUE)) {
  cat("\nFinal VIF values:\n")
  print(round(car::vif(final_model), 2))
}
```

## Visual: VIF Comparison

```{r, fig.width=8, fig.height=5}
if (requireNamespace("car", quietly = TRUE)) {
  # Compute VIF for original model
  model_full <- lm(species_richness ~ ., data = bioclim_example)
  vif_before <- car::vif(model_full)

  # VIF after modelPrune
  vif_after <- car::vif(final_model)

  # Combined barplot
  par(mar = c(8, 4, 4, 2))
  all_vars <- unique(c(names(vif_before), names(vif_after)))
  vif_combined <- data.frame(
    before = vif_before[match(all_vars, names(vif_before))],
    after = vif_after[match(all_vars, names(vif_after))]
  )
  vif_combined[is.na(vif_combined)] <- 0
  vif_combined <- vif_combined[order(vif_combined$before, decreasing = TRUE), ]

  # Show top 15
  n_show <- min(15, nrow(vif_combined))
  barplot(t(as.matrix(vif_combined[1:n_show, ])),
          beside = TRUE,
          las = 2,
          main = "VIF Before and After modelPrune()",
          ylab = "VIF",
          col = c(rgb(0.8, 0.2, 0.2, 0.7), rgb(0.2, 0.5, 0.8, 0.7)),
          cex.names = 0.6,
          names.arg = rownames(vif_combined)[1:n_show])
  abline(h = 5, col = "black", lwd = 2, lty = 2)
  legend("topright",
         legend = c("Before", "After", "Limit = 5"),
         fill = c(rgb(0.8, 0.2, 0.2, 0.7), rgb(0.2, 0.5, 0.8, 0.7), NA),
         border = c("black", "black", NA),
         lty = c(NA, NA, 2),
         lwd = c(NA, NA, 2),
         bty = "n")
}
```

## Comparison Summary

Compared to manual VIF removal, modelPrune():

- Automated: No manual iteration required
- Optimized: Uses graph algorithms to find optimal subset
- force_in support: Protect important variables from removal
- Transparent: Clear documentation of removed variables
- Consistent: Always returns same result

---

# Overview of Approaches

## Tool Comparison

| Your Goal | Tool |
|-----------|------|
| Remove redundant predictors (no response) | corrPrune() |
| Reduce VIF in regression model | modelPrune() |
| Maximize prediction accuracy | glmnet or Boruta |
| Mixed-type data (numeric + categorical) | assocSelect() |
| Protect certain variables from removal | corrselect (force_in) |
| Reproducible multicollinearity removal | corrselect |
| Quick exploratory analysis | caret::findCorrelation() |

## Design Characteristics

**corrselect**:

- Reproducible multicollinearity removal
- Maximal variable retention under correlation constraint (exact mode)
- Model-agnostic preprocessing
- Mixed-type data (numeric, factors, ordered)
- Protected variables (force_in)
- Unbiased coefficient estimates (no shrinkage)

**Alternatives**:

- **caret**: Quick exploration; reproducibility not required
- **Boruta**: Feature importance screening; not redundancy removal
- **glmnet**: Prediction-focused; coefficient shrinkage
- **Manual VIF**: Educational contexts; full manual control

## Combined Approach Example

```{r, eval=FALSE}
# Example workflow combining multiple approaches
data_cleaned <- corrPrune(raw_data, threshold = 0.7)          # Remove redundancy
model_data <- modelPrune(response ~ ., data_cleaned, limit = 5)  # Ensure VIF < 5
final_model <- lm(response ~ ., data = model_data)           # Interpretable model
```

---

## See Also

- `vignette("quickstart")` - Quick introduction to corrselect
- `vignette("workflows")` - Real-world workflow examples
- `vignette("advanced")` - Advanced features and customization
- `vignette("theory")` - Theoretical foundations and formulation
